---
title: "FDS_PS2_Q1_Group26"
author: "Group 26"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r}
rm(list=ls())
library(tidyverse)
library(ggplot2)
```

```{r}
table <- read.csv("projectB.csv")
```

# Strategy

```{r}
# calculate the weight of every stock
table1 <- table %>%
  group_by(tradedate) %>%
    mutate(avg_return = mean(return)) %>% 
  mutate(weight = -(return - avg_return)/690) %>%
  group_by(tradedate) %>%
  mutate(day_sum = sum(weight)) %>%  # check the sum of individual weights add up to 0 
  group_by(id) %>% 
  mutate(prev_weight = lag(weight))

table1$tradedate <- as.Date(table1$tradedate, format = "%m/%d/%Y")
```

```{r}
# calculate individual daily return and portfolio daily return 
table2 <- table1 %>%
  mutate(daily_ret = return * prev_weight) %>%
  group_by(tradedate) %>%
  summarise(portfolio_ret = sum(daily_ret)) 
```

```{r}
# calculate average market return  
table3 <- table1 %>% 
  group_by(tradedate) %>%
  summarise(avg_return = mean(return))
```

# Question 1: Strategy Simulation

## a) Plot and label a graph with the time series of daily portfolio returns pi(t) and another graph with the time series of daily market returns R(t)

```{r}
# daily portfolio returns 
table2 %>%
  ggplot(aes(x = tradedate, y = portfolio_ret)) +
  geom_line(aes(color=portfolio_ret)) + 
  labs(title = 'Time Series of Daily Portfolio Returns',
       x = 'Time',
       y = 'Daily Portfolio Returns')
```

```{r}
# daily portfolio returns 
table3 %>%
  ggplot(aes(x = tradedate, y = avg_return)) +
  geom_line(aes(color=avg_return)) + 
  labs(title = 'Time Series of Daily Market Returns',
       x = 'Time',
       y = 'Daily Market Returns')
```

## b) Annualized mean return, volatility, and Sharpe ratio of the strategy and of the market average

```{r}
table2 <- table2 %>% drop_na()
risk_free_rate <- 0
```

```{r}
# Strategy Stats
sd_strategy <- sd(table2$portfolio_ret)
annualized_mean_return_strategy <- 100*(mean(table2$portfolio_ret) * 252); annualized_mean_return_strategy
annualized_volatility_strategy <- 100*(sd_strategy * sqrt(252)); annualized_volatility_strategy

strategy_sharpe <- (annualized_mean_return_strategy - risk_free_rate) / (annualized_volatility_strategy); strategy_sharpe
```

```{r}
# Market Average Stats
sd_market_average <- sd(table3$avg_return)
annualized_mean_return_market <- 100*(mean(table3$avg_return) * 252); annualized_mean_return_market
annualized_volatility_market <- 100*(sd_market_average * sqrt(252)); annualized_volatility_market

market_sharpe <- (annualized_mean_return_market - risk_free_rate) / (annualized_volatility_market); market_sharpe
```

## c) Are they consistent over time? Is the strategy's distribution of daily returns stationary? 

```{r}
library(tseries)
adf.test(table2$portfolio_ret)
```

I conducted Augmented Dickey-Fuller Test, with a significance level of 0.05, to test for stationarity for the strategy daily returns, and they are consistent over time. The strategy's daily returns are subject to fluctuations in the market, but overall it is stationary and the volatility of the strategy is smaller than the volatility of the market. 

## d) Are there any events (market days) that are "unusual events"? Are there any stocks that are "outliers"? If so, do they have a siginificant effect on the overall strategy return? 

```{r}
# boxplot of market average return 
boxplot(table3$avg_return)
```

```{r}
# boxplot of portfolio return 
boxplot(table2$portfolio_ret)
```

```{r}
# account for unusual days with unexpected returns 
z_scores <- scale(table3$avg_return)
outliers <- abs(z_scores) > 2 
outlier_indices <- which(outliers, arr.ind = TRUE)
outlier_dates <- table2$tradedate[outlier_indices[, 1]]
outlier_dates
```


```{r}
# account for stocks 
individual_return <- table1 %>% 
  mutate(daily_return = return * prev_weight) %>%
  group_by(id) %>% 
  mutate(weight_all = sum(weight)) %>% 
  drop_na() %>% 
  group_by(id) %>% 
  mutate(return_all = sum(daily_return))
```

```{r}
individual_return_weight_top5 <- individual_return %>%
  group_by(id) %>% 
  arrange(desc(weight_all)) %>% 
  select('id','weight_all','return_all') %>% 
  distinct() %>% 
  head(5)
individual_return_weight_top5
```

```{r}
individual_return_weight_bottom5 <- individual_return %>%
  group_by(id) %>% 
  arrange(weight_all) %>% 
  select('id','weight_all','return_all') %>% 
  distinct() %>% 
  head(5)
individual_return_weight_bottom5
```

```{r}
individual_return_return_top5 <- individual_return %>%
  group_by(id) %>% 
  arrange(desc(return_all)) %>% 
  select('id','weight_all','return_all') %>% 
  distinct() %>% 
  head(5)
individual_return_return_top5
```

```{r}
individual_return_return_bottom5 <- individual_return %>%
  group_by(id) %>% 
  arrange(return_all) %>% 
  select('id','weight_all','return_all') %>% 
  distinct() %>% 
  head(5)
individual_return_return_bottom5
```


## e) What is the correlation between the strategy returns and the market return? Is the strategy in fact market-neutral or dollar neutral? 

```{r}
table3 <- table3[-1,]
correlation_strategy_market <- cor(table2$portfolio_ret, table3$avg_return); correlation_strategy_market
```

The correlation between the strategy returns and the market returns is 0.08774, suggesting that the strategy is not strongly correlated with the overall market.

## f) Do the strategy returns show autocorrelation? What does that say about the strategy?

```{r}
# Calculate the autocorrelation function
acf_result <- acf(table2$portfolio_ret, lag.max = 20); acf_result
```

Autocorrelation is a statistical measure that assesses the degree to which a time series is correlated with itself at different time lags. If the returns of the strategy exhibit significant autocorrelation, it indicates that there is some pattern or persistence in the returns over time. This may suggest that the strategy is not entirely random and that there may be some internal patterns that we can follow to improve the strategy. 

Here's a brief interpretation of these results:
- Positive autocorrelation (values greater than 0) at lag 1 suggests that there is a weak positive correlation between returns in consecutive periods.
- At lag 3, there is a relatively stronger positive autocorrelation of 0.038.
- There are some negative autocorrelations (values less than 0), particularly at lags 4, 5, 6, and 12, indicating that returns at these lags are negatively correlated with returns at lag 0 (the current period).
The autocorrelation coefficients vary in magnitude and sign at different lags, which suggests that there may be some degree of serial correlation in the returns, but it is not very strong or consistent.
- Overall, these results suggest that there might be some autocorrelation in the returns of the time series  especially at lag 1 and lag 3, but the autocorrelations are not overwhelmingly strong. 

## g) What is the maximum drawdown? The maximum drawdown is defined as the largest maginitude peak-to-trough decline in the cumulative return of the portfolio.

```{r}
cumulative_returns <- cumprod(1 + table2$portfolio_ret) - 1

time_periods <- 1:length(cumulative_returns)

# Plot the cumulative returns
plot(time_periods, cumulative_returns, type = "l", 
     xlab = "Time Period [1996-2000]", ylab = "Cumulative Returns", 
     main = "Cumulative Returns Over Time")
```

```{r}
max_drawdown <- max(diff(cumulative_returns - cummax(cumulative_returns))); max_drawdown
```

Maximum drawdown of the strategy is 0.04244

```{r}
cumulative_returns_market <- cumprod(1 + table3$avg_return) - 1

time_periods_market <- 1:length(cumulative_returns_market)

# Plot the cumulative returns
plot(time_periods_market, cumulative_returns_market, type = "l", 
     xlab = "Time Periods [1996-2000]", ylab = "Cumulative Returns", 
     main = "Cumulative Returns Over Time")
```


```{r}
max_drawdown_market <- max(diff(cumulative_returns_market - cummax(cumulative_returns_market))); max_drawdown_market
```


## h) What is the correlation between the long and short sub-portfolios? The long and short sub-portfolios consist on each day of those positions whose weights are positiev or negative

```{r}
long_portfolio <- table1 %>% 
  filter(weight > 0) %>% 
  group_by(tradedate) %>%
  summarise(avg_return_long = mean(return))

short_portfolio <- table1 %>% 
  filter(weight <= 0) %>% 
  group_by(tradedate) %>%
  summarise(avg_return_short = mean(return))
```

```{r}
correlation_long_short <- cor(long_portfolio$avg_return_long, short_portfolio$avg_return_short); correlation_long_short
```

The correlation between long and short sub-portfolios is 0.6021. 

## i) How realistic is this simulation? How implementable? Suppose that there were no transaction costs and no market impact. How closely do you think actual returns on this strategy would track the simulated values π(t)? List data issues that need to be considered and what their impact might be.

The contrarian strategy is theoretically realistic, but we also need to consider these factors in practically implementing the strategy. First of all,markets are not perfectly inefficient, and information about stocks is quickly incorporated into prices. Therefore, the assumption of excess returns persisting over time may not hold as strongly as assumed in the simulation. Moreover, market sentiment also factors into the decision-making process of the contrarian strategy. The success of contrarian strategies relies on the presence of behavioral biases among market participants, such as overreaction and herding behavior. The degree of these biases can vary over time and across markets.

In reality, the actual returns should follow the trend we predicted with the simulated values π(t), but deviations from the predictions should also be expected, particularly when the market undergoes larger than expected turbulence and fluctuations. In addition, the timing of trade executions can also impact the maginitude of deviation. For example, the contratian strategy relies on trading as late in the day as possible. In practice, this may not always be feasible, and trading may occur at suboptimal times. Real-world execution may not also be as precise as assumed in the simulation, and slippage occurs when we can't buy or sell at the exact closing price.

Data Issues that needs to be considered 
- Survivorship Bias: The universe contains 690 stocks that remain active from 1996 to 2000. In fact, considering the simulation I conducted, it should have accounetd for companies that have gone bankrupt or delisted in four years. Ignoring these can bias the results.
- Look-Ahead Bias: Using future information in simulations can lead to overly optimistic results. Simulations should only use data available at the time of trading decisions.
- Stationarity Assumption: The strategy's success depends on the stationarity of market conditions. Markets can change over time, and historical patterns may not persist, as suggested by the empirical counterexamples of the random walk hypothesis. 



